---
title: "CF"
author: "A.DeMartin"
date: "2025-06-13"
output: 
  html_document:
    keep_md: true
    toc: true
editor_options: 
  chunk_output_type: inline
---

```{r setup, echo = FALSE}
options(width = 100)
knitr::opts_chunk$set(warning = FALSE, message = FALSE, dev = c("png", "pdf"))
seed <- 1234
```

```{r}
library(here)
library(tximeta)
library(DESeq2)
library(vsn)
library(pheatmap)
library(RColorBrewer)
library(biomaRt)
library(tidyverse)
library(msigdbr)
library(clusterProfiler)
library(ggpubr)
library(EnhancedVolcano)
```

```{r}
basedir <- "/Users/immbio/Desktop/Project/Angelina/NC_HTx_03/data/"
```

## load data and create a SummarizedExperiment
```{r, eval=TRUE, include=TRUE}
target_folder <- file.path(basedir, "Ref/quants/CF")
folders <- list.dirs(path = target_folder, full.names = TRUE, recursive = FALSE)
folder_names <- basename(folders)

coldata <- data.frame(names = folder_names)

coldata$files <- file.path(paste0(folders), "quant.sf")

file.exists(coldata$files)
se <- tximeta(coldata)

dim(se)

head(rownames(se))
gse <- summarizeToGene(se)

dim(gse)
head(rownames(gse))

gse@assays
assayNames(gse)

coldata$names
```

### add metadata
```{r add meta, eval=TRUE, include=TRUE}
#assign stimulation condition
colData(gse)$stim <- "stim"
colData(gse)$stim[grepl("BMP4_50_24h", colData(gse)$names)] <- "BMP_50"
colData(gse)$stim[grepl("BMP4_500_24h", colData(gse)$names)] <- "BMP_500"
colData(gse)$stim[grepl("HGF_50_24h", colData(gse)$names)] <- "HGF_50"
colData(gse)$stim[grepl("HGF_500_24h", colData(gse)$names)] <- "HGF_500"
colData(gse)$stim[grepl("MFAP5_50_24h", colData(gse)$names)] <- "MFAP5_50"
colData(gse)$stim[grepl("MFAP5_200_24h", colData(gse)$names)] <- "MFAP5_200"
colData(gse)$stim[grepl("SB431542", colData(gse)$names)] <- "SB431542"
colData(gse)$stim[grepl("FCS_24h", colData(gse)$names)] <- "FCS"
colData(gse)$stim[grepl("BSA_24h", colData(gse)$names)] <- "BSA"
table(colData(gse)$stim)
colData(gse)$stim <- as.factor(colData(gse)$stim)

colData(gse)$stim2 <- "stim2"
colData(gse)$stim2[grepl("FCS|BSA", colData(gse)$names)] <- "ctr"
colData(gse)$stim2[grepl("BMP", colData(gse)$names)] <- "BMP"
colData(gse)$stim2[grepl("HGF", colData(gse)$names)] <- "HGF"
colData(gse)$stim2[grepl("MFAP5", colData(gse)$names)] <- "MFAP5"
colData(gse)$stim2[grepl("SB431542", colData(gse)$names)] <- "SB431542"
table(colData(gse)$stim2)
colData(gse)$stim2 <- as.factor(colData(gse)$stim2)

#assign patient code
colData(gse)$pc <- "pc"
colData(gse)$pc[grepl("Fibro1", colData(gse)$names)] <- "Fibro1"
colData(gse)$pc[grepl("Fibro3", colData(gse)$names)] <- "Fibro3"
table(colData(gse)$pc)
colData(gse)$pc <- as.factor(colData(gse)$pc)

saveRDS(gse, file=paste0(basedir, "gse_stimhiPS.rds"))
```

```{r, eval=TRUE, include=TRUE}
gse <- readRDS(file=paste0(basedir, "gse_stimhiPS.rds"))
```

## sample distance
### construct a DESeqDataSet stim
```{r construct DESeqDataSet object stim2}
##construct a DESeqDataSet object
#dds <- DESeqDataSet(gse, design = ~ cell + dex)
dds <- DESeqDataSet(gse, design = ~ stim2)

## pre-filtering
nrow(dds)
smallestGroupSize <- 2
keep <- rowSums(counts(dds) >= 10) >= smallestGroupSize
dds <- dds[keep,]
nrow(dds)

## variance stabilizing transformation
lambda <- 10^seq(from = -1, to = 2, length = 1000)
cts <- matrix(rpois(1000*100, lambda), ncol = 100)
meanSdPlot(cts, ranks = FALSE)

log.cts.one <- log2(cts + 1)
meanSdPlot(log.cts.one, ranks = FALSE)

vsd <- vst(dds, blind = FALSE)
rld <- rlog(dds, blind = FALSE)

dds <- estimateSizeFactors(dds)

##RNA-seq counts, however, the expected variance grows with the mean
df <- bind_rows(
  as_data_frame(log2(counts(dds, normalized=TRUE)[, 1:2]+1)) %>%
         mutate(transformation = "log2(x + 1)"),
  as_data_frame(assay(vsd)[, 1:2]) %>% mutate(transformation = "vst"),
  as_data_frame(assay(rld)[, 1:2]) %>% mutate(transformation = "rlog"))
  
colnames(df)[1:2] <- c("x", "y")  

lvls <- c("log2(x + 1)", "vst", "rlog")
df$transformation <- factor(df$transformation, levels=lvls)

ggplot(df, aes(x = x, y = y)) + geom_hex(bins = 80) +
  coord_fixed() + facet_grid( . ~ transformation)  
```

### sample distance
```{r distance, include=FALSE, eval=TRUE, echo=TRUE}
sampleDists <- dist(t(assay(rld)))
sampleDists
```

### distance heatmap
```{r distance heatmap}
sampleDistMatrix <- as.matrix( sampleDists )
rownames(sampleDistMatrix) <- paste(rld$stim, rld$pc, sep = " - " )
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette(rev(brewer.pal(9, "BuPu")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = colors)

colors <- colorRampPalette(rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = colors)
```

### distance MDS
```{r vis dim red bulk}
colCond <- c("#213782","#822145","#f4a582","#8c510a","#B45B5C","#725663FF","#D53E4F","grey","lightgrey")
names(colCond) <- c("BMP_50", "BMP_500", "HGF_50", "HGF_500", "MFAP5_50", "MFAP5_200","SB431542","FCS", "BSA")

colpc <- c("#ABDDA4","#779d8d")
names(colpc) <- c("Fibro1","Fibro3")

#plotPCA(rld, intgroup = c("cond", "grp"))
plotPCA(rld, intgroup = c("stim", "pc")) +
  geom_point(aes(color = stim), size = 8, alpha = 0.8) +
  scale_color_manual(values = colCond) # size = dot size, alpha = transparency

mds <- as.data.frame(colData(rld))  %>%
         cbind(cmdscale(sampleDistMatrix))
ggplot(mds, aes(x = `1`, y = `2`, color = stim, shape = pc)) +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(values = colCond) +
  geom_point(size = 4) + coord_fixed() + ggtitle("MDS with rld data")
```

## BMP vs ctr
### diff expressed in BMP4 vs ctr
```{r}
dds <- DESeq(dds)
res <- results(dds, contrast = c("stim2", "BMP", "ctr")) ## define which conditions to contrast
#res <- results(dds)
res

genes <- data.frame(ensID=rownames(res))
mart <- useDataset("hsapiens_gene_ensembl", useMart("ensembl"))

G_list <- getBM(attributes=c('hgnc_symbol', 'ensembl_gene_id'), 
      filters = 'ensembl_gene_id', 
      values = genes$ensID, 
      mart = mart)

resDat <- data.frame(res@listData) %>% mutate(ensembl_gene_id = rownames(res)) %>% 
  left_join(., G_list, by= "ensembl_gene_id")
```

### top 50 heatmap
```{r top 50 BMP vs ctr heatmap, fig.height=12, fig.width=6}
colCond2 <- c("#355C7D","#f4a582","#B45B5C","#A59C94","#D53E4F")
names(colCond2) <- c("BMP","HGF","MFAP5","ctr", "SB431542")

topGenes <- head(order(res$padj),50)
#topGenes <- resDat$ensembl_gene_id[which(resDat$padj < 0.01)]

mat  <- assay(rld)[ topGenes, ]
namesDat <- data.frame(ensembl_gene_id = rownames(mat)) %>%
  left_join(., G_list, by= "ensembl_gene_id")

# Define annotation colors
ann_colors <- list(
  stim2 = colCond2,
  pc = colpc,
  stim = colCond)

anno <- as.data.frame(colData(rld)[, c("stim2", "stim" ,"pc")])

# custom heatmap colors
heat_colors <- colorRampPalette(c("#2166AC", "#F7F7F7", "#B2182B"))(50)

pheatmap(mat, annotation_col = anno, annotation_colors = ann_colors, scale = "row",
         labels_row = namesDat$hgnc_symbol, color = heat_colors, labels_col = rep("", ncol(mat)))
```

### top 100 heatmap
```{r top 100 BMP vs ctr heatmap, fig.height=18, fig.width=6}
topGenes <- head(order(res$padj),100)
#topGenes <- resDat$ensembl_gene_id[which(resDat$padj < 0.01)]

mat  <- assay(rld)[ topGenes, ]
namesDat <- data.frame(ensembl_gene_id = rownames(mat)) %>%
  left_join(., G_list, by= "ensembl_gene_id")

# Define annotation colors
ann_colors <- list(
  stim2 = colCond2,
  pc = colpc,
  stim = colCond)

anno <- as.data.frame(colData(rld)[, c("stim2", "stim" ,"pc")])

# custom heatmap colors
heat_colors <- colorRampPalette(c("#2166AC", "#F7F7F7", "#B2182B"))(50)

pheatmap(mat, annotation_col = anno, annotation_colors = ann_colors, scale = "row",
         labels_row = namesDat$hgnc_symbol, color = heat_colors, labels_col = rep("", ncol(mat)))
```

### vulcano
```{r vulcano BMP vs ctr, fig.height=10, fig.width=13}
res$symbol <- resDat$hgnc_symbol
EnhancedVolcano(res,
    lab = res$symbol,
    x = 'log2FoldChange',
    y = 'pvalue',
    xlim = c(-10, 10),
    #ylim = c(0,50),
    pCutoff = 10e-05,
    FCcutoff = 1,
    title = "DeSEQ2 results",
    subtitle = bquote(italic("BMP4 vs ctr")))
```

### pathway analysis
```{r pathway anaysis BMP vs ctr, fig.height=8, fig.width=10}
rankedGenes <- resDat %>%
  filter(!is.na(ensembl_gene_id)) %>%
  mutate(rank = log2FoldChange) %>%
  arrange(desc(rank)) %>%
  pull(rank, ensembl_gene_id)

head(rankedGenes)

rankedGenes <- resDat %>%
  filter(!is.na(ensembl_gene_id)) %>%
  mutate(rank = -log10({pvalue}) * sign({log2FoldChange})) %>%
  filter(!is.na(rank)) %>% 
  arrange(desc(rank)) %>%
  pull(rank, ensembl_gene_id)

head(rankedGenes)


term2gene <- msigdbr(species = "Homo sapiens", category = "H") %>% 
  dplyr::select(gs_name, ensembl_gene)
term2name <- msigdbr(species = "Homo sapiens", category = "H") %>% 
  dplyr::select(gs_name, gs_description) %>% 
  distinct()

gseaRes <- GSEA(rankedGenes,
                TERM2GENE = term2gene,
                TERM2NAME = term2name,
                pvalueCutoff = 1.00, 
                minGSSize = 5,
                maxGSSize = 500)

gseaResDat <- gseaRes@result

as_tibble(gseaRes) %>% 
  arrange(desc(abs(NES))) %>% 
  top_n(20, wt=-p.adjust) %>% 
  dplyr::select(-core_enrichment) %>%
  mutate(across(c("enrichmentScore", "NES"), round, digits=3)) %>% 
  mutate(across(c("pvalue", "p.adjust", "qvalue"), scales::scientific))
```

```{r pathway anaysis BMP vs ctr-2, fig.height=15, fig.width=10}
dotplot(gseaRes, showCategory=10, split=".sign") + facet_grid(.~.sign)
```

```{r pathway anaysis BMP vs ctr-3, fig.height=5, fig.width=10}
selGODat <- gseaResDat %>% 
  arrange(desc(abs(NES))) %>% 
  top_n(20, wt=-p.adjust) %>% 
  mutate(qscore=-log(p.adjust, base=10)) %>% 
  mutate(cluster="BMP")

p <- ggbarplot(selGODat, x = "ID", y = "qscore",
          fill = "cluster",               
          color = "cluster",            
          palette = colCond2,            
          sort.val = "asc",           
          sort.by.groups = TRUE      
          #x.text.angle = 90           
          ) + 
  rotate()
p
```

## HGF vs ctr
### diff expressed in HGF vs ctr
```{r}
dds <- DESeq(dds)
res <- results(dds, contrast = c("stim2", "HGF", "ctr")) ## define which conditions to contrast
#res <- results(dds)
res

genes <- data.frame(ensID=rownames(res))
mart <- useDataset("hsapiens_gene_ensembl", useMart("ensembl"))

G_list <- getBM(attributes=c('hgnc_symbol', 'ensembl_gene_id'), 
      filters = 'ensembl_gene_id', 
      values = genes$ensID, 
      mart = mart)

resDat <- data.frame(res@listData) %>% mutate(ensembl_gene_id = rownames(res)) %>% 
  left_join(., G_list, by= "ensembl_gene_id")
```

### top 50 heatmap
```{r top 50 HGF vs ctr heatmap, fig.height=12, fig.width=6}

topGenes <- head(order(res$padj),50)
#topGenes <- resDat$ensembl_gene_id[which(resDat$padj < 0.01)]

mat  <- assay(rld)[ topGenes, ]
namesDat <- data.frame(ensembl_gene_id = rownames(mat)) %>%
  left_join(., G_list, by= "ensembl_gene_id")

# Define annotation colors
ann_colors <- list(
  stim2 = colCond2,
  pc = colpc,
  stim = colCond)

anno <- as.data.frame(colData(rld)[, c("stim2", "stim" ,"pc")])

# custom heatmap colors
heat_colors <- colorRampPalette(c("#2166AC", "#F7F7F7", "#B2182B"))(50)

pheatmap(mat, annotation_col = anno, annotation_colors = ann_colors, scale = "row",
         labels_row = namesDat$hgnc_symbol, color = heat_colors, labels_col = rep("", ncol(mat)))
```

### top 100 heatmap
```{r top 100 HGF vs ctr heatmap, fig.height=18, fig.width=6}
topGenes <- head(order(res$padj),100)
#topGenes <- resDat$ensembl_gene_id[which(resDat$padj < 0.01)]

mat  <- assay(rld)[ topGenes, ]
namesDat <- data.frame(ensembl_gene_id = rownames(mat)) %>%
  left_join(., G_list, by= "ensembl_gene_id")

# Define annotation colors
ann_colors <- list(
  stim2 = colCond2,
  pc = colpc,
  stim = colCond)

anno <- as.data.frame(colData(rld)[, c("stim2", "stim" ,"pc")])

# custom heatmap colors
heat_colors <- colorRampPalette(c("#2166AC", "#F7F7F7", "#B2182B"))(50)

pheatmap(mat, annotation_col = anno, annotation_colors = ann_colors, scale = "row",
         labels_row = namesDat$hgnc_symbol, color = heat_colors, labels_col = rep("", ncol(mat)))
```

### vulcano
```{r vulcano HGF vs ctr, fig.height=10, fig.width=13}
res$symbol <- resDat$hgnc_symbol
EnhancedVolcano(res,
    lab = res$symbol,
    x = 'log2FoldChange',
    y = 'pvalue',
    xlim = c(-10, 10),
    ylim = c(0,20),
    pCutoff = 10e-05,
    FCcutoff = 1,
    title = "DeSEQ2 results",
    subtitle = bquote(italic("HGF vs ctr")))
```

## MFAF5 vs ctr
### diff expressed in MFAP5 vs ctr
```{r}
dds <- DESeq(dds)
res <- results(dds, contrast = c("stim2", "MFAP5", "ctr")) ## define which conditions to contrast
#res <- results(dds)
res

genes <- data.frame(ensID=rownames(res))
mart <- useDataset("hsapiens_gene_ensembl", useMart("ensembl"))

G_list <- getBM(attributes=c('hgnc_symbol', 'ensembl_gene_id'), 
      filters = 'ensembl_gene_id', 
      values = genes$ensID, 
      mart = mart)

resDat <- data.frame(res@listData) %>% mutate(ensembl_gene_id = rownames(res)) %>% 
  left_join(., G_list, by= "ensembl_gene_id")
```

### top 50 heatmap
```{r top 50 MFAP5 vs ctr heatmap, fig.height=12, fig.width=6}
topGenes <- head(order(res$padj),50)
#topGenes <- resDat$ensembl_gene_id[which(resDat$padj < 0.01)]

mat  <- assay(rld)[ topGenes, ]
namesDat <- data.frame(ensembl_gene_id = rownames(mat)) %>%
  left_join(., G_list, by= "ensembl_gene_id")

# Define annotation colors
ann_colors <- list(
  stim2 = colCond2,
  pc = colpc,
  stim = colCond)

anno <- as.data.frame(colData(rld)[, c("stim2", "stim" ,"pc")])

# custom heatmap colors
heat_colors <- colorRampPalette(c("#2166AC", "#F7F7F7", "#B2182B"))(50)

pheatmap(mat, annotation_col = anno, annotation_colors = ann_colors, scale = "row",
         labels_row = namesDat$hgnc_symbol, color = heat_colors, labels_col = rep("", ncol(mat)))
```

### top 100 heatmap
```{r top 100 MFAP5 vs ctr heatmap, fig.height=18, fig.width=6}
topGenes <- head(order(res$padj),100)
#topGenes <- resDat$ensembl_gene_id[which(resDat$padj < 0.01)]

mat  <- assay(rld)[ topGenes, ]
namesDat <- data.frame(ensembl_gene_id = rownames(mat)) %>%
  left_join(., G_list, by= "ensembl_gene_id")

# Define annotation colors
ann_colors <- list(
  stim2 = colCond2,
  pc = colpc,
  stim = colCond)

anno <- as.data.frame(colData(rld)[, c("stim2", "stim" ,"pc")])

# custom heatmap colors
heat_colors <- colorRampPalette(c("#2166AC", "#F7F7F7", "#B2182B"))(50)

pheatmap(mat, annotation_col = anno, annotation_colors = ann_colors, scale = "row",
         labels_row = namesDat$hgnc_symbol, color = heat_colors, labels_col = rep("", ncol(mat)))
```

### vulcano
```{r vulcano MFAP5 vs ctr, fig.height=10, fig.width=13}
res$symbol <- resDat$hgnc_symbol
resdf <- as.data.frame(res)
EnhancedVolcano(res,
    lab = res$symbol,
    x = 'log2FoldChange',
    y = 'pvalue',
    xlim = c(-10, 10),
    ylim = c(0,12),
    pCutoff = 10e-05,
    FCcutoff = 1,
    title = "DeSEQ2 results",
    subtitle = bquote(italic("MFAP5 vs ctr")))
```

## SB431542 vs ctr
### diff expressed in SB431542 vs ctr
```{r}
dds <- DESeq(dds)
res <- results(dds, contrast = c("stim2", "SB431542", "ctr")) ## define which conditions to contrast
#res <- results(dds)
res

genes <- data.frame(ensID=rownames(res))
mart <- useDataset("hsapiens_gene_ensembl", useMart("ensembl"))

G_list <- getBM(attributes=c('hgnc_symbol', 'ensembl_gene_id'), 
      filters = 'ensembl_gene_id', 
      values = genes$ensID, 
      mart = mart)

resDat <- data.frame(res@listData) %>% mutate(ensembl_gene_id = rownames(res)) %>% 
  left_join(., G_list, by= "ensembl_gene_id")
```

### top 50 heatmap
```{r top 50 SB431542 vs ctr heatmap, fig.height=12, fig.width=6}
topGenes <- head(order(res$padj),50)
#topGenes <- resDat$ensembl_gene_id[which(resDat$padj < 0.01)]

mat  <- assay(rld)[ topGenes, ]
namesDat <- data.frame(ensembl_gene_id = rownames(mat)) %>%
  left_join(., G_list, by= "ensembl_gene_id")

# Define annotation colors
ann_colors <- list(
  stim2 = colCond2,
  pc = colpc,
  stim = colCond)

anno <- as.data.frame(colData(rld)[, c("stim2", "stim" ,"pc")])

# custom heatmap colors
heat_colors <- colorRampPalette(c("#2166AC", "#F7F7F7", "#B2182B"))(50)

pheatmap(mat, annotation_col = anno, annotation_colors = ann_colors, scale = "row",
         labels_row = namesDat$hgnc_symbol, color = heat_colors, labels_col = rep("", ncol(mat)))
```

### top 100 heatmap
```{r top 100 SB431542 vs ctr heatmap, fig.height=18, fig.width=6}
topGenes <- head(order(res$padj),100)
#topGenes <- resDat$ensembl_gene_id[which(resDat$padj < 0.01)]

mat  <- assay(rld)[ topGenes, ]
namesDat <- data.frame(ensembl_gene_id = rownames(mat)) %>%
  left_join(., G_list, by= "ensembl_gene_id")

# Define annotation colors
ann_colors <- list(
  stim2 = colCond2,
  pc = colpc,
  stim = colCond)

anno <- as.data.frame(colData(rld)[, c("stim2", "stim" ,"pc")])

# custom heatmap colors
heat_colors <- colorRampPalette(c("#2166AC", "#F7F7F7", "#B2182B"))(50)

pheatmap(mat, annotation_col = anno, annotation_colors = ann_colors, scale = "row",
         labels_row = namesDat$hgnc_symbol, color = heat_colors, labels_col = rep("", ncol(mat)))
```

### vulcano
```{r vulcano SB431542 vs ctr, fig.height=10, fig.width=13}
res$symbol <- resDat$hgnc_symbol
resdf <- as.data.frame(res)
EnhancedVolcano(res,
    lab = res$symbol,
    x = 'log2FoldChange',
    y = 'pvalue',
    xlim = c(-10, 10),
    #ylim = c(0,12),
    pCutoff = 10e-05,
    FCcutoff = 1,
    title = "DeSEQ2 results",
    subtitle = bquote(italic("SB431542 vs ctr")))
```

## BMP_50 vs BMP_500
### construct a DESeqDataSet stim
```{r construct DESeqDataSet object stim}
##construct a DESeqDataSet object
#dds <- DESeqDataSet(gse, design = ~ cell + dex)
dds <- DESeqDataSet(gse, design = ~ stim)

## pre-filtering
nrow(dds)
smallestGroupSize <- 2
keep <- rowSums(counts(dds) >= 10) >= smallestGroupSize
dds <- dds[keep,]
nrow(dds)

## variance stabilizing transformation
lambda <- 10^seq(from = -1, to = 2, length = 1000)
cts <- matrix(rpois(1000*100, lambda), ncol = 100)
meanSdPlot(cts, ranks = FALSE)

log.cts.one <- log2(cts + 1)
meanSdPlot(log.cts.one, ranks = FALSE)

vsd <- vst(dds, blind = FALSE)
rld <- rlog(dds, blind = FALSE)

dds <- estimateSizeFactors(dds)

##RNA-seq counts, however, the expected variance grows with the mean
df <- bind_rows(
  as_data_frame(log2(counts(dds, normalized=TRUE)[, 1:2]+1)) %>%
         mutate(transformation = "log2(x + 1)"),
  as_data_frame(assay(vsd)[, 1:2]) %>% mutate(transformation = "vst"),
  as_data_frame(assay(rld)[, 1:2]) %>% mutate(transformation = "rlog"))
  
colnames(df)[1:2] <- c("x", "y")  

lvls <- c("log2(x + 1)", "vst", "rlog")
df$transformation <- factor(df$transformation, levels=lvls)

ggplot(df, aes(x = x, y = y)) + geom_hex(bins = 80) +
  coord_fixed() + facet_grid( . ~ transformation)  
```

### diff expressed in BMP_50 vs BMP_500
```{r}
dds <- DESeq(dds)
res <- results(dds, contrast = c("stim", "BMP_50", "BMP_500")) ## define which conditions to contrast
#res <- results(dds)
res

genes <- data.frame(ensID=rownames(res))
mart <- useDataset("hsapiens_gene_ensembl", useMart("ensembl"))

G_list <- getBM(attributes=c('hgnc_symbol', 'ensembl_gene_id'), 
      filters = 'ensembl_gene_id', 
      values = genes$ensID, 
      mart = mart)

resDat <- data.frame(res@listData) %>% mutate(ensembl_gene_id = rownames(res)) %>% 
  left_join(., G_list, by= "ensembl_gene_id")
```

### top 50 heatmap
```{r top 50 BMP_50 vs BMP_500 heatmap, fig.height=12, fig.width=6}
topGenes <- head(order(res$padj),50)
#topGenes <- resDat$ensembl_gene_id[which(resDat$padj < 0.01)]

mat  <- assay(rld)[ topGenes, ]
namesDat <- data.frame(ensembl_gene_id = rownames(mat)) %>%
  left_join(., G_list, by= "ensembl_gene_id")

# Define annotation colors
ann_colors <- list(
  stim2 = colCond2,
  pc = colpc,
  stim = colCond)

anno <- as.data.frame(colData(rld)[, c("stim2", "stim" ,"pc")])

# custom heatmap colors
heat_colors <- colorRampPalette(c("#2166AC", "#F7F7F7", "#B2182B"))(50)

pheatmap(mat, annotation_col = anno, annotation_colors = ann_colors, scale = "row",
         labels_row = namesDat$hgnc_symbol, color = heat_colors, labels_col = rep("", ncol(mat)))
```

### vulcano
```{r vulcano BMP_50 vs BMP_500, fig.height=10, fig.width=13}
res$symbol <- resDat$hgnc_symbol
EnhancedVolcano(res,
    lab = res$symbol,
    x = 'log2FoldChange',
    y = 'pvalue',
    xlim = c(-10, 10),
    ylim = c(0,10),
    pCutoff = 10e-05,
    FCcutoff = 1,
    title = "DeSEQ2 results",
    subtitle = bquote(italic("BMP_50 vs BMP_500")))
```

## plot selected genes
### indiv counts
```{r sel genes, fig.height=4, fig.width=6}
colCond2 <- c("#355C7D","#A59C94","#f4a582","#B45B5C","#D53E4F")

sel_genes <- c("ENSG00000125378", "ENSG00000107779", "ENSG00000138696", "ENSG00000180875", "ENSG00000105976", "ENSG00000197614", "ENSG00000019991", "ENSG00000276886")

for(ID in sel_genes){
  symbol <- G_list$hgnc_symbol[which(G_list$ensembl_gene_id == ID)]
  plotCounts(dds, gene = ID, intgroup=c("stim2"), main = symbol, col=colCond2[factor(dds$stim2)],
             pch = 19)
}
```

## session info
```{r session info}
sessionInfo()
date()
```
